---
title: "Effect of Supplement and Dose of supplement on Guinea Pig Tooth Growth"
author: "Juan Bosco Mendoza Vega"
date: "September 18 2016"
output:
  word_document: default
  pdf_document:
    fig_height: 3
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(comment = "")
knitr::opts_chunk$set(strip.white = T)
```

```{r, message = F}
library(dplyr)
library(car)
```

## Overview

A brief summary of the data set *ToothGrowth* will be presented and means of the effect of Supplement type and Dose of supplement in the mean *Tooth Growth* of guinea pigs will be tested. 
*t-Tests* will be conducted, using confidence intervals and significance levels to determine differences in the means of *Tooth Growth*.
 
## Summary of the data set

There data set contains three variables and sixty observations. According to the summary and documentation called by `?ToothGrowth`, the variables are *Tooth length* (len, numeric); *Supplement type* (supp, factor with two levels. OJ: Orange Juice, VC: Ascorbic acid); and *Dose* in miligrams per day (dose, numeric with 3 values. 0.5, 1 and 2) (Appendix A). 

At a glance, there seems to be differences in *Tooth Growth* by *Supplement*, and *Tooth Growth* seems to increase as *Dose* increases (Appendix B).

## Assumptions

After conducting a *Shapiro-Wilks test* (Appendix C) and visually exploring *Tooth length* (Appendix X), this variable seems to be **normally distributed**. The results of conducting a *Levene's test* for homogeneity of variance (Appendix D) show evidence that the groups to compare **don't have equal variance**. Also, there's no evidence that the observations are independent, so it's assumed they are **non paired  groups**.

## Choice of test

Considering our stated assumptions, a ***t-Test* for non paired groups with unequal variance will be conducted**. In all cases, our null hypothesis ($H_0$) is that means of the groups are equal. When comparing by *Supplement*, the alternative hypthesis ($H_a$) is that means of the groups are different; when comparing by *Dose*, the alternative hypothesis is that the mean of the first group is less than the mean of the second (code for these tests is found in Appendix E).

## Comparison by Supplement type

$H_a$: Orange juice $\neq$ Ascorbic acid
```{r t-Test supp}
t.test(
    formula = len ~ supp,
    data = ToothGrowth,
    var.equal = F, 
    paired = F, 
    alternative = "two.sided"
)
```
The 95% confidence interval contains 0 and the significance level larger than 0.05, so there are no differences in the *Tooth Growth* mean by *Supplement*.

## Comparison by Dose

$H_a$: 0.5 miligrams per day $<$ 1 miligram per day
```{r .5 v 1}
dose_0.5 <- ToothGrowth %>% filter(dose == 0.5) %>% select(len)
dose_1.0   <- ToothGrowth %>% filter(dose == 1)   %>% select(len)
dose_2.0   <- ToothGrowth %>% filter(dose == 2)   %>% select(len)

t.test(
    x = dose_0.5,
    y = dose_1.0,
    var.equal = F, 
    paired = F, 
    alternative = "less"
)
```
The 95% confidence interval is entirely below 0 and the significance level is less than 0.05. A dose of 0.5 miligrams has a lower *Tooth Growth* mean than a dose of 1 miligram.

$H_a$: 0.5 miligram per day $<$ 2 miligram per day
```{r .5 v 2}
t.test(
    x = dose_0.5,
    y = dose_2.0,
    var.equal = F, 
    paired = F, 
    alternative = "less"
)
```
The 95% confidence interval is entirely below 0 and the significance level is less than 0.05. A dose of 0.5 miligrams has a lower *Tooth Growth* mean than a dose of 2 miligrams.


$H_a$: 1 miligram per day $<$ 2 miligram per day
```{r 1 v 2}
t.test(
    x = dose_1.0,
    y = dose_2.0,
    var.equal = F, 
    paired = F, 
    alternative = "less"
)
```
The 95% confidence interval is entirely below 0 and the significance level is less than 0.05. A dose of 1 miligram has a lower *Tooth Growth* mean than a dose of 2 miligrams.

## Conclusion

With this results, we can conclude that there is no statistically significant effect of Supplement on the mean of *Tooth Growth*. In contrast, there are statistically significant effects of the Dose on the mean of *Tooth Growth*. As the *Dose* increases, the mean of *Tooth Growth* also increases.

# Appendix

## A. Structure of the **ToothGrowth** data set.

```{r}
data("ToothGrowth")
str(ToothGrowth)
summary(ToothGrowth)
```

## B. Visualization of the groups

```{r boxplots, fig.align='left'}
par(mfrow = c(1, 2))
boxplot(formula = len ~ supp, data = ToothGrowth, 
    col = c("#FF6666", "#66BBFF"),
    xlab = "Supplement type", ylab = "Tooth length")

boxplot(formula = len ~ dose, data = ToothGrowth, 
    col = c("#FFFF66"),
    xlab = "Dose in miligrams per day", ylab = "Tooth length")
```

## C. Test of normality of Tooth length
The ¨*Shapiro-Wilk Normality Test* is used, The null hypothesis for this test is that the given data is normally distributed. Results indicate we can't reject this hypotesis, so **we assume the distribution is normal**.
```{r normality, echo = T}
shapiro.test(ToothGrowth$len)
```

A histogram with an overlaid density curve of this variable and a Normal Q-Q plot are generated to support the assumtion its distribution is normal.
```{r histogram qqplot, echo = T}
par(mfrow = c(1, 2))
hist(ToothGrowth$len, 
    probability = T, ylim = c(0, 0.05), 
    main = "Histogram and Density\n of Tooth length", xlab = "Tooth length",
    col = "#FFFF66")
lines(density(ToothGrowth$len))

qqnorm(y = ToothGrowth$len, 
    main = "Normal Q-Q Plot\n for Tooth length",
    pch = 21, bg = c("#FFFF66"))
qqline(y = ToothGrowth$len)
```

## D. Test if variance is equal across groups
The *Levene's test of homogenity of variance is used*. Groups are: *Supplement type*; *Dose in miligrams per day*; and *Supplement type* and *Dose in miligrams per day*. Results indicate groups **don't have equal variance**.

```{r levene by supp, echo = T}
library(car)
leveneTest(len ~ supp, data = ToothGrowth)
leveneTest(len ~ as.factor(dose), data = ToothGrowth)
```

## E. Code for t-Tests
```{r, echo = T, eval = F}
library(dplyr)

t.test(formula = len ~ supp, data = ToothGrowth,
    var.equal = F, paired = F, alternative = "two.sided")

dose_0.5 <- ToothGrowth %>% filter(dose == 0.5) %>% select(len)
dose_1.0 <- ToothGrowth %>% filter(dose == 1)   %>% select(len)
dose_2.0 <- ToothGrowth %>% filter(dose == 2)   %>% select(len)

t.test(x = dose_0.5, y = dose_1.0,
    var.equal = F, paired = F, alternative = "less")
t.test(x = dose_0.5, y = dose_2.0,
    var.equal = F, paired = F, alternative = "less")
t.test(x = dose_1.0, y = dose_2.0,
    var.equal = F, paired = F, alternative = "less")
```
    
